---
title: "Governance Portfolio Review"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 10,
  dpi = 300
)

library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(ggthemes)

devtools::load_all()

theme_set(theme_minimal())

wb_projects_gov <- portfolioreview::wb_projects_gov

wb_lending_gov <- wb_projects_gov |>
  filter(product_line_type == "Lending Product")

wb_asa_gov <- wb_projects_gov |>
  filter(product_line_type == "Analytic and Advisory Activities Product")

theme_labels <- function(x) {
  case_match(
    x,
    "pfm"                  ~ "Public Financial Management",
    "procurement"          ~ "Public Procurement",
    "public_admin"         ~ "Public Administration",
    "environmental_social" ~ "Environmental & Social"
  )
}
```

## Overview

The dataset covers active World Bank operations led by the Governance Global Practice (GOV) in IDA-eligible and Blend countries, approved between FY2019 and FY2026. Data were extracted from the [World Bank Data Explorer](https://dataexplorer.worldbank.org/) between April 21–29, 2026. The dataset includes both lending operations and analytic and advisory activities (ASAs), with a thematic focus on governance-related issues. 

The analysis below provides an overview of lending and ASA trends over time, as well as a thematic breakdown of projects in the portfolio. It also provides regional comparisons in the lending and ASA portoflios.

## Lending trends

```{r lending-trends}
wb_lending_gov |>
  summarise(rate = n_distinct(proj_id), .by = proj_approval_fy) |>
  ggplot(aes(proj_approval_fy, rate)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  )
```

## Lending trends by region

```{r lending-trends-region}
wb_lending_gov |>
  summarise(rate = n_distinct(proj_id), .by = c(proj_approval_fy, region)) |>
  ggplot(aes(proj_approval_fy, rate, color = region)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  scale_color_solarized() +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  ) +
  facet_wrap(
    vars(region),
    ncol = 2
  ) +
  theme(legend.position = "bottom")
```

## ASA trends

```{r asa-trends}
wb_asa_gov |>
  summarise(rate = n_distinct(proj_id), .by = proj_approval_fy) |>
  ggplot(aes(proj_approval_fy, rate)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  )
```

## ASA trends by region

```{r asa-trends-region}
wb_asa_gov |>
  summarise(rate = n_distinct(proj_id), .by = c(proj_approval_fy, region)) |>
  ggplot(aes(proj_approval_fy, rate, color = region)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  scale_color_solarized() +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  ) +
  theme(legend.position = "bottom")
```

## Thematic breakdown over time

```{r theme-trends}
wb_lending_gov |>
  summarise(
    rate_pfm                  = sum(theme_pfm),
    rate_procurement          = sum(theme_procurement),
    rate_public_admin         = sum(theme_public_admin),
    rate_environmental_social = sum(theme_env_social),
    .by = proj_approval_fy
  ) |>
  pivot_longer(
    cols      = starts_with("rate_"),
    names_to  = "theme_category",
    values_to = "rate"
  ) |>
  mutate(theme_category = theme_labels(str_remove(theme_category, "rate_"))) |>
  ggplot(aes(proj_approval_fy, rate, color = theme_category)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  scale_color_solarized() +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  ) +
  theme(legend.position = "bottom")
```

## Thematic breakdown over time by region

```{r theme-trends-region}
wb_lending_gov |>
  summarise(
    rate_pfm                  = sum(theme_pfm),
    rate_procurement          = sum(theme_procurement),
    rate_public_admin         = sum(theme_public_admin),
    rate_environmental_social = sum(theme_env_social),
    .by = c(proj_approval_fy, region)
  ) |>
  pivot_longer(
    cols      = starts_with("rate_"),
    names_to  = "theme_category",
    values_to = "rate"
  ) |>
  mutate(theme_category = theme_labels(str_remove(theme_category, "rate_"))) |>
  ggplot(aes(proj_approval_fy, rate, color = theme_category)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  scale_color_solarized() +
  facet_wrap(vars(region), ncol = 2) +
  labs(
    x = "Fiscal year of project approval",
    y = "Number of projects"
  ) +
  theme(legend.position = "bottom")
```

## Thematic breakdown in FY2026

```{r theme-breakdown-2026}
wb_lending_gov |>
  filter(proj_approval_fy == 2026) |>
  summarise(
    rate_pfm                  = sum(theme_pfm),
    rate_procurement          = sum(theme_procurement),
    rate_public_admin         = sum(theme_public_admin),
    rate_environmental_social = sum(theme_env_social)
  ) |>
  pivot_longer(
    cols      = starts_with("rate_"),
    names_to  = "theme_category",
    values_to = "rate"
  ) |>
  mutate(theme_category = theme_labels(str_remove(theme_category, "rate_"))) |>
  ggplot(aes(theme_category, rate, fill = theme_category)) +
  geom_col(alpha = 0.7) +
  scale_fill_solarized() +
  labs(
    x = "Theme",
    y = "Number of projects"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 30, hjust = 1)
  )
```

## Thematic breakdown in FY2026 by region

```{r theme-breakdown-2026-region}
wb_lending_gov |>
  filter(proj_approval_fy == 2026) |>
  summarise(
    rate_pfm                  = sum(theme_pfm),
    rate_procurement          = sum(theme_procurement),
    rate_public_admin         = sum(theme_public_admin),
    rate_environmental_social = sum(theme_env_social),
    .by = region
  ) |>
  pivot_longer(
    cols      = starts_with("rate_"),
    names_to  = "theme_category",
    values_to = "rate"
  ) |>
  mutate(theme_category = theme_labels(str_remove(theme_category, "rate_"))) |>
  ggplot(aes(theme_category, rate, fill = theme_category)) +
  geom_col(alpha = 0.7) +
  scale_fill_solarized() +
  facet_wrap(vars(region), ncol = 2) +
  labs(
    x = "Theme",
    y = "Number of projects"
  ) +
  theme(legend.position = "none")
```
